---
title: "ICAM1 Associations"
author: "Wanjun Gu"
format: pdf
editor: visual
---

```{r}
suppressPackageStartupMessages({
  library(dplyr)
  library(data.table)
  library(ggplot2)
  library(ggpubr)
})
geno = fread("genotype_master.csv")
pheno = fread("cdp_phenotyping_master.csv")

icam1 = select(geno, id, ICAM1_rs1799969G)

icam1 = left_join(pheno, icam1, by = "id") |>
  filter(genotype_exclude == 0)
```

*From the merged dataset, we select key variables for modeling: **SBP (`sbp_mean`)**, the ICAM1 genotype (`ICAM1_rs1799969G`, coded additively as 0/1/2 for A/G alleles), **sex**, and **BMI**.*

```{r}
icam1 = icam1 |>
  select(sbp_mean, ICAM1_rs1799969G, sex, bmi)
```

### *Linear Regression: Additive Model (without BMI)*

*We first assess the additive effect of the G allele on SBP, adjusting only for sex.*

```{r}
lm(data = icam1, formula = sbp_mean ~ ICAM1_rs1799969G + sex) |> 
  summary()
```

### *Linear Regression: Additive Model (with BMI)*

*Next, we adjust for potential confounding by body mass index (BMI).*

```{r}
lm(data = icam1, formula = sbp_mean ~ ICAM1_rs1799969G + sex + bmi) |> 
  summary()
```

*In both models above, the **G allele of rs1799969** is significantly associated with **higher systolic blood pressure**, suggesting a potential functional role of this variant in regulating vascular physiology.*

### *Linear Regression: Genotype as Categorical (without BMI)*

*We now treat genotype as a categorical variable to allow for potential non-linear effects.*

```{r}
lm(data = icam1, formula = sbp_mean ~ as.factor(ICAM1_rs1799969G) + sex) |> 
  summary()
```

### *Linear Regression: Genotype as Categorical (with BMI)*

*We repeat this model, now adjusting for BMI.*

```{r}
lm(data = icam1, formula = sbp_mean ~ as.factor(ICAM1_rs1799969G) + sex + bmi) |> 
  summary()
```

*These models reinforce our previous findings. Individuals carrying the **G allele** tend to have **higher SBP**, even when accounting for sex and BMI. This effect is consistent whether the genotype is treated as a continuous dosage or as a factor.*

### *Visualization*

*To visualize the relationship between genotype and SBP, we create a boxplot stratified by genotype and colored by sex.*

```{r}
plt = ggplot(data = icam1, aes(x = as.factor(ICAM1_rs1799969G), y = sbp_mean)) + 
  geom_boxplot(outlier.shape = NA) + 
  geom_point(aes(fill = sex), size = 2.5, color = "black", shape = 21, position = position_dodge2(0.2)) + 
  scale_fill_manual(values = c("#80cba4", "#4965b0")) + 
  xlab("ICAM1 rs1799969 (A>G)") + 
  ylab("Systolic Blood Pressure (mmHg)") + 
  labs(fill = "") +
  theme_linedraw() + 
  theme(text = element_text(size = 15))

suppressWarnings(print(plt))
```

```{r}
ggsave(filename = "icam1_association.png", device = "png", dpi = 1200, height = 5, width = 4)
```

### *Interpretation*

*These results indicate that **rs1799969 in the ICAM1 gene is significantly associated with systolic blood pressure levels**, where **carriers of the G allele tend to exhibit higher SBP**.*

*This association suggests a **protective role of the A allele**, which may contribute to lower vascular pressure through mechanisms involving endothelial adhesion or immune signaling, given ICAM1’s known biological functions.*

*Interestingly, population genetic data show that the **A allele is substantially more common in Andean populations** compared to other global populations. This observation, combined with our association analysis, suggests that the **Andean-enriched A allele is associated with lower blood pressure**, potentially representing an adaptive cardiovascular trait in high-altitude environments.*

*Further studies are warranted to investigate the functional mechanisms underlying this association and to explore whether similar effects are observed in other cohorts and ancestries.*
